Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-09-04 manu-draft.Rmd
editing discussion

Legacy wastewater discharges from Piney Point did not occur again until 2021, although onsite stormwater management and discharges have occurred throughout its history.  Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains. The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.  Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged at Berth 12 in Port Manatee.  This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.  From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.  Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep03a10].

@@

The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point together represent an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations provide a weight of evidence that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.  These anomalous events included 1) a large, spring diatom bloom in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for an unusual amount of observations supported by the decades of long-term monitoring data as a baseline for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).  Although these out-of-range results had specific temporal and spatial patterns during the period of observation, they were seen throughout the bay suggesting a system-wide effect of effluent from Piney Point.  In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.  Spill events reported to FDEP have been documented for years (e.g., industrial spills, stormwater pump failures, sanitary sewer overflows) and spill volumes for these for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point. Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point discharged over the ten day period exceeded current annual estimates of the external load into lower Tampa Bay.  From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in the Bay in 2021 is an outlier compared to past years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator and asset for environmental quality.

Several of the correlations in Figure \@ref(fig:allcor) support known responses of water quality to nutrient loads and can be explained through mechanistic relationships among water quality, macroalgae, and seagrasses.  Regarding the nutrient management paradigm for Tampa Bay, the linear association between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations and loadings from the PCA in Figure \@ref(fig:allpca). An increase in nitrogen was associated with an increase in chlorophyll, which in turn was associated with a decrease in water clarity.  Temporally, these changes were observed at different times and by different species of phytoplankton.  The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa.  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen already utilized by different taxa, including macroalgae.

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability that may negatively affect seagrass growth.  However, changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, area 3 closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient (as shown in the PCA in Figure \@ref(fig:allpca)) and not any association with effluent from Piney Point.  Correlations among the macroalgae and seagrasses can also be explained by spatial differences among species along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Seagrasses are clonal organisms that grow together and species are usually not randomly dispersed in the same patch.  Shoal grass and turtle grass typically grow at shallower depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels, blades of shoal grass and turtle grass can "lay down" during low tide to minimize dessication stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species is better suited at deeper depths that are minimally affected by varying tidal height.  As such, the correlation analysis can be explained by spatial partitioning of seagrasses along depth gradients due to adaptive characteristics of each species.  These associations may also reflect natural variation in water clarity along the longitidinal axis of Tampa Bay.  Shoal grass can occur in more turbid waters that increase towards the northern portion of Tampa Bay, which are also shallower.

Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria sp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva sp.*, *Caulerpa sp.*) and seasonally occur earlier in the growing season.  The dominance of the red groups early in the period of record followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.  However past spill events were associated with an increase in both red and green macroalgae in smaller subembayments of Tampa Bay, prompting manual removal to prevent conditions associated with hypoxia from decomposition of macroalgal biomass [@Switzer11]. Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites as a result of nutrient cycling stimulated by wastewater release. A clear difference in the macroalgae community in 2021 compared to past years was a high abundance of cyanobacteria macroalgae (i.e., Lyngbya sp.).  Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee was observed in 2021. The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.  *Lyngbya sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.  Although the seagrass community did not change significantly across the study period, long-term impacts may be observed.  Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year and half in southwest Florida.  These changes may have resulted from a reduction in water clarity from heavy concentrations of *K. brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.  Species shifts from macroalgal  to seagrass dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality of *Caulerpa* dominance relative to seagrasses.  General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communites.  Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Managagement District.  From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest drivers acting at a regional scale are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.  Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
* Risk of decline (IRL ex.), regression of past progress

>>>>

Legacy wastewater discharges from Piney Point did not occur again until 2021, although onsite stormwater management and discharges have occurred throughout its history.  Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains. The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.  Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged at Berth 12 in Port Manatee.  This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.  From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.  Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

@@

The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay in June and July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).  Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.

In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.  Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay. An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point. Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of the external load into lower Tampa Bay [@tbep0417].  From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in 2021 is an outlier compared to past years. These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.

Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)) and loadings from the PCA results (Figure \@ref(fig:allpca)). Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability that may negatively affect seagrass growth.  However, changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient [Figure \@ref(fig:allpca), @Sherwood17].  Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper. These associations may also reflect natural variation in water clarity along the main axis of Tampa Bay.  Shoal grass can occur in more turbid waters of oligohaline areas.

Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria sp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva sp.*, *Caulerpa sp.*) and seasonally occur earlier in the growing season.  The dominance of the red groups early in the period of record followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.  However past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11]. Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release. A clear difference in the macroalgae community in 2021 compared to past years was a high abundance of cyanobacteria macroalgae (i.e., *Lyngbya sp.*).  Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee was observed in 2021. The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.  *Lyngbya sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.  Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red time bloom in 2018.  Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality.  General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].  Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Management District.  From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.  Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

Fisheries resources have undoubtedly been impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in abundance and diversity of fish species in Tampa Bay in 2021 are forthcoming. Routine sampling by the FWC Fisheries Independent Monitoring Program occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].

Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.  For benthic communities, sediments were sampled in April during the discharge near Port Manatee and surrounding waters.  These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106]. Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions. Results suggested a mix of conditions dominated by "fair" and "good" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].  Comparison of the initial 2021 samples to historical benthic conditions suggested relatively consistent sediment quality from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).  Additional sampling is planned later at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.  Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021. This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass loss for southwest Florida.

Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly to lower Tampa Bay instead of through channels to back-bay habitats (i.e., Bishop Harbor).  The comparison to past years provided a stark contrast of the effect of wastewater discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impacts on different habitats that varies spatially and throughout the water column.  The Tampa Bay Coastal Ocean Model [@Chen19] was used to simulate daily tracer particle movement and expected dilution of nutrient concentrations from the point of discharge.  Although these results were initially used to inform monitoring locations, the model suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient for phytoplankton production. Although simulated particle tracking showed that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of inorganic nitrogen from stack effluent (Table \@ref(tab:stktab)) and the background concentrations for the bay (Table \@ref(tab:wqsumtab)).

In the broader context of mining impacts to surface waters, the results herein reinforce the understanding that legacy pollutants from phosphate mining represent a serious threat to environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18].  The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed likely related to low pH of the effluent.  Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  A common theme is that legacy wastewater from fertilizer production is poorly maintained and long-term plans are insufficient to dispose of pollutants that can significantly affect coastal resources.  These are not isolated examples and sustained attention and regulatory oversight on efforts to close these facilities in an environmentally responsible manner are needed at the local, national and international level.

Limitations of our analysis are also important to note to inform future event-based monitoring and additional research to fully understand acute, short-term threats to the sustained recovery of Tampa Bay.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  The weight of evidence that 2021 is an anomaly relative to past years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more definite description of the recycling and fate of nitrogen from Piney Point.  There may likely never be a definitive link, but this is not necessary to support future actions that ensure the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.  These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
